Exploring the internal language of toposes

### 2018-07-03 slides-como2018.tex
Work around bug in the PDF specification

@@

  Beautiful constructive proofs can also be found in Richman's note on
  \href{https://www.ams.org/journals/proc/1988-103-04/S0002-9939-1988-0954974-5/S0002-9939-1988-0954974-5.pdf}{nontrivial
  uses of nontrivial rings} and in the
  \href{https://arxiv.org/abs/1605.04832}{recent textbook by Lombardi and
  Quitté} on constructive commutative algebra.}

@@

  Using the internal language of toposes, we will give a short, conceptual and
  constructive proof of Grothendieck's generic freeness lemma. Again, if so
  desired, one can unwind the internal proof to obtain a
  constructive proof which doesn't reference topos theory. The proof obtained
  in this way is still an improvement on the previously known proofs, requiring
  no advanced prerequisites in commutative algebra, and takes
  \href{https://rawgit.com/iblech/internal-methods/master/paper-generic-freeness.pdf}{about
  a page}.}

@@

  The standard internal language of toposes in not enough for our purposes, as
  it misses unbounded quantification (``for all groups'', ``for all rings'')
  and dependent types. Shulman's \href{https://arxiv.org/abs/1004.3802}{stack
  semantics} offers what we need. No knowledge of stacks is necessary to enjoy
  his paper. Prior work includes
  \href{https://www.cl.cam.ac.uk/~amp12/papers/polist/polist.pdf}{Polymorphism
  is Set Theoretic, Constructively} by Pitts and
  \href{http://www.phil.cmu.edu/projects/ast/Papers/Awodey-Butz-Simpson-Streicher-APAL-2013.pdf}{Relating
  first-order set theories, toposes and categories of classes} by Awodey, Butz,
  Simpson and Streicher (obtained independently and published after Shulman).}

@@

  Constructively, there are several non-equivalent forms of the intermediate
  value theorem. The following version doesn't admit an intuitionistic proof:
  \begin{indentblock}Let~$g : \RR \to \RR$ be a function between the (Dedekind)
  reals which is continuous in the usual epsilon/delta sense. Assume~$g(-1) < 0
  < g(1)$. Then there exists a number~$x \in \RR$ such that~$g(x) = 0$.
  \end{indentblock}
  If there was an intuitionistic proof, the statement would hold in any topos,
  so in particular in sheaf toposes over topological spaces. By the
  translations shown on the slide (the corner quotes ``$\speak{\ldots}$''
  indicate that translation into formal language is left to the reader), this
  would amount to the following strengthening of the intermediate value
  theorem: In continuous families of continuous functions, zeros can locally be
  picked continuously. However, this strengthening is invalid, as
  \href{https://raw.githubusercontent.com/iblech/internal-methods/master/images/zeros-in-families.mp4}{this
  video shows}.

@@

  \jnote{3}{One can show, assuming that the little Zariski topos is
  \emph{overt}, that the module~$M$ in~$\Set$ and the module~$\ull{M}$ of the
  little Zariski topos share all first-order properties. This observation
  explains the metatheorem displayed at the bottom left. The assumption is
  satisfied if any element of~$A$ is nilpotent or not nilpotent, so it's always
  satisfied if the law of excluded middle is available. In an intuitionistic
  context, it's still ``morally satisfied''. Details are in Section~12.9 of
  \href{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
  notes}.}

>>>>

% Workaround for the issue described at
% https://tex.stackexchange.com/questions/164406/beamer-using-href-in-notes.
\newcommand{\fixedhref}[2]{\makebox[0pt][l]{\hspace*{\paperwidth}\href{#1}{#2}}\href{#1}{#2}}

@@

  Beautiful constructive proofs can also be found in Richman's note on
  \fixedhref{https://www.ams.org/journals/proc/1988-103-04/S0002-9939-1988-0954974-5/S0002-9939-1988-0954974-5.pdf}{nontrivial
  uses of nontrivial rings} and in the
  \fixedhref{https://arxiv.org/abs/1605.04832}{recent textbook by Lombardi and
  Quitté} on constructive commutative algebra.}

@@

  Using the internal language of toposes, we will give a short, conceptual and
  constructive proof of Grothendieck's generic freeness lemma. Again, if so
  desired, one can unwind the internal proof to obtain a
  constructive proof which doesn't reference topos theory. The proof obtained
  in this way is still an improvement on the previously known proofs, requiring
  no advanced prerequisites in commutative algebra, and takes
  \fixedhref{https://rawgit.com/iblech/internal-methods/master/paper-generic-freeness.pdf}{about
  a page}.}

@@

  The standard internal language of toposes in not enough for our purposes, as
  it misses unbounded quantification (``for all groups'', ``for all rings'')
  and dependent types. Shulman's \fixedhref{https://arxiv.org/abs/1004.3802}{stack
  semantics} offers what we need. No knowledge of stacks is necessary to enjoy
  his paper. Prior work includes
  \fixedhref{https://www.cl.cam.ac.uk/~amp12/papers/polist/polist.pdf}{Polymorphism
  is Set Theoretic, Constructively} by Pitts and
  \fixedhref{http://www.phil.cmu.edu/projects/ast/Papers/Awodey-Butz-Simpson-Streicher-APAL-2013.pdf}{Relating
  first-order set theories, toposes and categories of classes} by Awodey, Butz,
  Simpson and Streicher (obtained independently and published after Shulman).}

@@

  Constructively, there are several non-equivalent forms of the intermediate
  value theorem. The following version doesn't admit an intuitionistic proof:
  \begin{indentblock}Let~$g : \RR \to \RR$ be a function between the (Dedekind)
  reals which is continuous in the usual epsilon/delta sense. Assume~$g(-1) < 0
  < g(1)$. Then there exists a number~$x \in \RR$ such that~$g(x) = 0$.
  \end{indentblock}
  If there was an intuitionistic proof, the statement would hold in any topos,
  so in particular in sheaf toposes over topological spaces. By the
  translations shown on the slide (the corner quotes ``$\speak{\ldots}$''
  indicate that translation into formal language is left to the reader), this
  would amount to the following strengthening of the intermediate value
  theorem: In continuous families of continuous functions, zeros can locally be
  picked continuously. However, this strengthening is invalid, as
  \fixedhref{https://raw.githubusercontent.com/iblech/internal-methods/master/images/zeros-in-families.mp4}{this
  video shows}.

@@

  \jnote{3}{One can show, assuming that the little Zariski topos is
  \emph{overt}, that the module~$M$ in~$\Set$ and the module~$\ull{M}$ of the
  little Zariski topos share all first-order properties. This observation
  explains the metatheorem displayed at the bottom left. The assumption is
  satisfied if any element of~$A$ is nilpotent or not nilpotent, so it's always
  satisfied if the law of excluded middle is available. In an intuitionistic
  context, it's still ``morally satisfied''. Details are in Section~12.9 of
  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
  notes}.}
